### Activity 1.3 Solutions ###

Solutions to the exercises in Activity 1.3.

#### Exercise 1.3.1: Numeric Vectors
1) Create a vector from 3 to 11, in steps of 2

2) Create another vector from 12 to 4 in steps of -2

3) Display the last 2 elements of the second vector

4) Calculate the mean of the element-wise product of these
two vectors

##### Solutions 
First vector:
```{r}
x <- seq(3, 11, by = 2)
```

Second vector:
```{r}
y <- seq(12, 4, by = -2)
y
```

Determine the length of the vector:
```{r}
n <- length(y)
```

Select the last two elements:
```{r}
y[c(n - 1, n)]
```

Calculate the mean of the element-wise product of x and y:
```{r}
mean(x * y)
```


#### Exercise 1.3.2: Character vectors 

1) Create a character vector consisting of nouns (of the
length of your choosing)

2) Create a second character vector of same length with verbs

3) Paste the two vectors together

4) Count the total number of characters in the entire noun
vector.  Hint, use the 'sum' function


##### Solutions

Create the noun vector:
```{r}
nouns <- c("cats", "dogs", "elephants")
```

Create the verb vector:
```{r}
verbs <- c("bark", "climb", "jump")
```

Join them together:
```{r}
paste(nouns, verbs)
```

Count the number of characters:
```{r}
sum(nchar(nouns))
```


#### Exercise 1.3.3: Logical vectors
1) Create random numeric vector of length 1000 with values
uniformly chosen between 0 and 1.  Hint: use the `runif`
function

2) Count the number of elements in the vector that
fall in the interval `[0.2, 0.8]`.  On average, it should
be 600


##### Solutions

Generate 1000 random uniform draws in the interval `[0, 1]`.
```{r}
x <- runif(1000)
```

Create the indicator for >= 0.2.
```{r}
above_2 <- x >= 0.2
```

Create the indicator for <= 0.8.
```{r}
below_8 <- x <= 0.8
```

Count the number of 'TRUE's that satisfy both conditions.
```{r}
sum(above_2 & below_8)
```

#### Exercise 1.3.5: Named vectors 


1) Create a named character vector of length 3

2) Create a second named logical vector of length 3

3) Switch the names of the 2 vectors


##### Solutions

A named character vector:
```{r}
xChar <- c(this = "one", that = "too", these = "many")
xChar
```

A named logical vector:
```{r}
xLogical <- c(happy = TRUE, sad = FALSE, angry = TRUE)
xLogical
```

Switch the names.  Begin by storing the names of `xChar` in a character vector.
```{r}
xCharNames <- names(xChar)
```

Assign the names of `xLogical` to `xChar`.
```{r}
names(xChar) <- names(xLogical)
```

Assign the names of `xChar` to `xLogical`.
```{r}
names(xLogical) <- xCharNames
```

Look at the results
```{r}
xChar
xLogical
```

#### Exercise 1.3.6: Data frames

1) Create a data frame with at least 7 rows with the
following columns:
    1. A character vector consisting of some repetition of
       the letters 'a', 'b' and 'c'
    2. A numeric vector of randomly generated numbers between
       0 and 1.  Hint:  use runif()
    3. A second numeric vector of randomly generated numbers
       between 2 and 7.  Hint: use runif() again

 2) Display the number of rows of the data frame

 3) Assign rownames consisting of capital letters, 'A', 'B',
    'C', etc.  Hint, use the `LETTERS` object that is part of
    R

 4) Display the second column of the fourth and fifth rows
    using the row and column names of the data frame

 5) Create a subset of the data by selecting those observations
    meet all of the following criteria:
    a.  The character vector has a value of 'b' or 'c'
    b.  The value of the first numeric vector is greater than
        0.1
    c.  The value of the second numeric vector is less than
        6.


##### Solutions 

Create the data frame.  I'm using vectors of length 9.
```{r}
X <- data.frame(char = rep(letters[1:3], 3),
                num1 = runif(9),
                num2 = runif(9, min = 2, max = 7))
```

Get the number of rows of the data frame.
```{r}
n <- nrow(X)
n
```

Assign rownames:
```{r}
rownames(X) <- LETTERS[1:n]
```

Display the data frame.
```{r}
X
```

Display 2nd column, 4th and 5th rows using names.
```{r}
X[c("D", "E"), "num1"]
```


Create a subset of the data to satisfy the criteria.
```{r}
criteria1 <- X$char %in% c("b", "c")
criteria2 <- X$num1 > 0.1
criteria3 <- X$num2 < 6
allCrit <- criteria1 & criteria2 & criteria3
```

How many rows satisfy the criteria?
```{r}
sum(allCrit)
```

Display the subset:
```{r}
X[allCrit,]
```

What happens if all values of `allCrit` are false? We get back a data frame with no rows
```{r}
allCrit <- rep(FALSE, n)
X[allCrit,]
```

#### Exercise 1.3.7: Matrices

 1) Create a 2 x 2 matrix with the following elements:
    (2, 4, 7, 1), loaded rowwise.

 2) Calculate the trace (the sum of the diagonal elements)
    of the matrix.  Hint: use the sum() function

 3) Create a column vector of length 2 with the elements
    (3, 5)

 4) Using matrix multiplication, and your results from (1)
    and (3), find the product of the transpose of the column
    vector, the matrix, and the column vector.  Display the
    result should as a single number (not a 1 x 1 matrix).


##### Solutions

Create the matrix:
```{r}
m <- matrix(c(2, 4, 7, 1), nrow = 2, byrow = TRUE)
m
```

Calculate the trace:
```{r}
sum(diag(m))
```

Create the column vector:

```{r}
v <- c(3, 5)
v
```

Calcuate the product, display as a vector:
```{r}
as.vector(t(v) %*% m %*% v)
```

#### Exercise 1.3.8:  Lists


 1) Create a named list with 3 elements:  a character vector,
    a numeric vector, and a 2 x 2 matrix

 2) Extract the third and first elements of the list,
    in that order

 3) Extract the 2nd element without the elment name

 4) Calculate the length of each element of the list and
    return the result as a vector


##### Solutions

Create a list with three named elements
```{r}
myList <- list(charVec = c("this", "that"),
               numVec = rnorm(7),
               mat = matrix(1:4, nrow = 2))
myList
```

Extract the third and first elements. There are two ways to do this:
```{r}
myList[c("mat", "charVec")]
myList[c(3, 1)]
```

Extract the 2nd element. Therea re two ways to do this:
```{r}
myList[["numVec"]]
myList$numVec
```

Calculate the length of each element, returned as a vector.
```{r}
unlist(lapply(myList, length))
```

