# Intro to R "Big Data" Bootcamp #
## Activity 1: Introduction to R ##

### Activity 1.2: Getting Started ###
========================================================
#### Activity 1.2.1: Help commands
When you don't know exactly what you're looking for,
search for a topic.  There are two ways to do this:

```{r eval=FALSE}
help.search("graphics")
??graphics
```

 When you know the name of the R function, e.g. the function
 print(), there are two ways to do this:

```{r eval=FALSE}
help(print)
?print
```

Sometimes you just need a reminder about the arguments (the inputs) of a function.  The args() function can help in
this situation.  Here are the arguments for the rnorm()
function, which generates normal random variates

```{r}
args(rnorm)
```
Examples are often included in the documentation of functions at the bottom of the documentation.  Look at:

```{r eval=FALSE}
?seq
```
 You can execute the example code for a function all at once
 using the example() function:

```{r}
example(seq)
```

#### Activity 1.2.2:Demos


 Some R packages include demonstrations you can run to learn
 about their capability.  Be sure to hit "Enter" to
 cycle through the components of each demo

 Here's one from the 'graphics' package, which gives a nice
 demonstration of the breadth of graphical capability in R

```{r}
demo("graphics")
```
 And here's one about generalized linear models (glm)

```{r}
demo(lm.glm)
```

#### Activity 1.2.3: R as a calculator


 Addition

```{r}
7 + 3
```
Subtraction

```{r}
7 - 3
```
Multiplication

```{r}
7 * 3
```
Division

```{r}
7 / 3
```
Integer division

```{r}
7 %/% 3
```
Division remainder (the modulo)

```{r}
7 %% 3
```
exponentiation

```{r}
exp(7)
```
Natural logarithm

```{r}
log(7)
```
Base 10 log

```{r}
log10(1000)
```
Examples of common mathematical functions and constants

```{r}
sqrt(16)
cos(pi)
```
Powers:  2 to the 5th

```{r}
2^5
```
Large and small numbers

```{r}
1.7e+05
1.7e-03
```
And there are numerous statistical functions.  For example:
The probability a standard normal random variable takes
a value of 1.96 or lower:

```{r}
pnorm(1.96)
```
And the probability of 4 heads in 12 fair coin flips

```{r}
dbinom(4, 12, 0.5)
```
And, best of all, it handles division by 0 gracefully :)

```{r}
1 / 0
```



